---
title: "Lab 2 Practice"
author: "Chelsea Sanford"
format:
  html:
    code-fold: show
    toc: true
    number-sections: true
editor: visual
excecute: 
  echo: true
  message: false
  warning: false
---

# Time Series Practice

## Libraries

```{r}
library(tidyverse)
library(here)
library(tsibble)
library(feasts)
library(fable)
```

## Load in the data

```{r}
maunaloa_df <- read.csv(here('data/co2_mauna_loa.csv'))
```

## Convert the dates and turn into a time series
#### Use Lubridate to convert the dates from character to DATE

```{r}
maunaloa_ts <- maunaloa_df %>% 
  mutate(date = tsibble::yearmonth(date)) %>% 
  as_tsibble(key = NULL,
             index = date)

ggplot(maunaloa_ts, aes(x = date, y = co2_mean)) +
  geom_line() + 
  labs(x = 'Date' , y = 'Mean CO2 Concentration (ppm)\nMauna Loa Station')
```

## Exploratory Subseries Plot: Filter by Date and Time
#### Use filter_index() to filter by year

```{r}
maunaloa_1980 <- maunaloa_ts %>% 
  filter_index('1980-01-01' ~ .)
```

#### Create a Subseries Plot
```{r}
maunaloa_1980 %>% 
  gg_subseries(y = co2_mean)
```

## Analysis Part 2

### Section 1

```{r}
dcmp <- maunaloa_ts %>% 
  model(feasts::STL(co2_mean ~ season(period = '1 year') +
                      trend(window = 49)))

components(dcmp) %>% 
  autoplot()
```


### Section 2: Create ETS Exponential Smoothing Model

```{r}
### Create a model
maunaloa_fit <- maunaloa_ts %>% 
  model(ets = ETS(co2_mean ~ season(method = "M") + trend(method = "A")))

maunaloa_forecast <- maunaloa_fit %>% 
  forecast(h = "20 years")

maunaloa_forecast %>% 
  autoplot(maunaloa_ts) +
  labs(x = 'Date', y = 'CO2 Mean')
```

### Section 3

```{r}
maunaloa_predicted <- maunaloa_fit %>% 
  broom::augment()

ggplot(maunaloa_predicted) +
  geom_line(aes(x = date, y = co2_mean)) +
  geom_line(aes(x = date, y = .fitted), color = 'red', alpha = .7)

ggplot(maunaloa_predicted, aes(x = .resid)) +
  geom_histogram()
```

### Section 4

```{r}
maunaloa_fit_03 <- maunaloa_ts %>% 
  filter_index(. ~ '2003 Jan') %>% 
  model(ets = ETS(co2_mean ~ season(method = "M") + trend(method = "A")))

maunaloa_forecast <- maunaloa_fit_03 %>% 
  forecast(h = "20 years")

maunaloa_forecast %>% 
  autoplot(maunaloa_ts) +
  labs(x = 'Date', y = 'CO2 Mean')
```

## Follow up Questions:

#### 1. There is yearly seasonality, where [co2] is high in spring and low in fall. The trend is increasing. There is no pattern in the noise.

#### 2. In your ETS time series model, explain whether an additive or multiplicative model would make more sense, in terms of the relationship between carbon dioxide, humans, and environmental factors.

The trend is additive because the data increases continuously. The seasonality is multiplicative because the data changes variability and has non-linear trends.This is due to environmental factors, while the upward trend is based on anthropoegenic factors.









